\documentclass{article}

\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{semantic-markup}
\usepackage{octave}
\usepackage{fancyvrb}
\frenchspacing
\usepackage[
    pdftitle={The octave Package for Typesetting Musical Pitches with Octave
    Designations},
    pdfauthor={Andrew A. Cashner},
    pdfsubject={LaTeX package},
    pdfkeywords={LaTeX, music, musicology, humanities}
]{hyperref}

\title{The \code{octave} Package for Typesetting Musical Pitches with Octave
Designations}
\author{Andrew A. Cashner\thanks{%
    \href{mailto:andrewacashner@gmail.com}
    {\nolinkurl{andrewacashner@gmail.com}}%
    }%
}

\begin{document}
\maketitle

There are two standard ways of indicating the octave of a musical pitch:
\begin{enumerate}
    \item The Helmholtz system, which uses subscript numerals after the pitch
        letter name, where \soCalled{middle C} is \pitch{C}{4}
    \item {\octaveprimes
        The traditional system, which uses upper- and lowercase letters followed
        by prime (tick) marks, where \soCalled{middle C} is \pitch{C}{4} and
        an octave higher is \pitch{C}{5}.}
\end{enumerate}

This package allows you to typeset pitch names conveniently using either method
(or even switching between) without worrying about typography, with a simple,
semantically meaningful interface.
Authors can write their documents without worrying about which system of
nomenclature their editor or press will prefer.
Simply changing the option with which the package is called will change the
display of all the pitches in the document that have been written with this
package's commands.

\tableofcontents 

\section{Package Options}

\begin{center}
    \begin{tabular}{ll}
        \code{number}   & (Default) Use Helmholtz numbers \\
        \code{prime}    & Use the traditional system with prime/tick marks \\
    \end{tabular}
\end{center}

To use octave numbers, just add \verb|\usepackage{octave}| to your preamble. 
To use the traditional system, write \verb|\usepackage[prime]{octave}| instead.



\section{Switching Systems within One Document}

At any time, you can change from one system to the other.

\begin{center}
\begin{tabular}{ll}
    \verb|\octaveprimes|    
        & Switch to the traditional system (prime/tick marks) \\
    \verb|\octavenumbers|   
        & Switch to the Helmholtz system (numbers) \\
\end{tabular}
\end{center}

\section{Style}

The default setup (stored in the command \verb|\pitchfont|) is for pitch letters
and numbers in the Helmholtz system to be in the normal font, while pitch
letters and prime marks are in italic.  You can change this, for example:

\verb|\renewcommand{\pitchfont}{\mdseries}|.

\section{Entering Pitch Names}

To enter a pitch name, use the \verb|\pitch| command.
It takes three arguments:
\begin{enumerate}
    \item (Mandatory) Letter name of the pitch: Can be upper- or lowercase; the
        package will adjust the case as necessary
    \item (Optional, in square brackets) Accidental command, e.g., \verb|$\sharp$| or \verb|\sh|
        from the \code{semantic-markup} package
    \item (Mandatory) Number of the pitch in the Helmholtz system
\end{enumerate}

Yes, the package requires you to use the numbered system, even if you want the
output to be in the traditional system. 
Thus middle C is \verb|\pitch{C}{4}|, and an octave and a semitone above that is
\verb|\pitch{C}[\sh]{5}| (using the \code{semantic-markup} package for the
accidental).


\section{Table of Octave Designations in the Two Systems}

The package also provides the \verb|\octavetable| which displays a comparison
of the two systems, shown in table~\ref{octavetable}.

\begin{table}
    \caption{Pitch designations in traditional (\code{prime}) and Helmholtz
    (\code{number}) systems}
    \label{octavetable}
    \begin{center}
    \octavetable
    \end{center}
\end{table}

\section{Code}

\VerbatimInput{octave.sty}

\end{document}
